# Literature review example

Let's use `ellmer` to pull structured data out of a PDF — a small taste of what a programmatic literature review looks like. This is an active area: LLMs have been tested for extracting ecological information from papers [@gougherty_testing_2024] and for synthesising information on potential pest controllers [@scheepens_large_2024], both with mixed but promising accuracy. I'll try summarising my own [paper on turtle fishing](https://conbio.onlinelibrary.wiley.com/doi/10.1111/conl.13056).

```{r eval=TRUE, echo=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

```{r}
x <- content_pdf_url("https://conbio.onlinelibrary.wiley.com/doi/epdf/10.1111/conl.13056")
```

This fails with a 403 — the server is blocking the request, correctly guessing I'm calling the PDF programmatically. Downloading the PDF manually and reading it locally works fine instead. Save the PDF into a `pdf-examples/` folder in your project, then point at it:

```{r}
mypdf <- content_pdf_file("pdf-examples/Brown_etal2024 national scale turtle mortality.pdf")
```

Now set up a chat and ask for structured data using `ellmer`'s `type_object()` helpers:

```{r}
cheap_model <- "anthropic/claude-3.5-haiku"

chat <- chat_openrouter(
    system_prompt = "You are a research assistant who specialises in extracting structured data from scientific papers.",
    model = cheap_model,
    api_args = list(max_tokens = 1000)
)

paper_stats <- type_object(
    sample_size = type_number("Sample size of the study"),
    year_of_study = type_number("Year data was collected"),
    method = type_string("Summary of statistical method, one paragraph max")
)

turtle_study <- chat$chat_structured(mypdf, type = paper_stats)
turtle_study$sample_size
turtle_study$year_of_study
turtle_study$method
```

It works — but check it carefully. In my case the "sample size" it returned was actually a model output (an estimated number of turtles caught), not the survey's real sample size. This paper reports several different sample sizes depending on which method you're looking at, so the prompt needs sharpening if you want it consistently right, especially before batch-processing many papers.

## Batch processing

Wrap the extraction in a function and apply it across several abstracts. Save each abstract as a `.txt` file in the same `pdf-examples/` folder:

```{r}
process_abstract <- function(file_path, chat) {
    abstract_text <- readLines(file_path, warn = FALSE)
    chat$chat_structured(abstract_text, type = paper_stats)
}
```

To reduce hallucinated answers when the abstract doesn't contain the information, mark fields optional:

```{r}
paper_stats <- type_object(
    sample_size = type_number("Number of surveys conducted to estimate turtle catch", required = FALSE),
    turtles_caught = type_number("Estimate for number of turtles caught", required = FALSE),
    year_of_study = type_number("Year data was collected", required = FALSE),
    region = type_string("Country or geographic region of the study", required = FALSE)
)

abstract_files <- list.files(path = "pdf-examples", pattern = "\\.txt$", full.names = TRUE)
results <- lapply(abstract_files, function(file) process_abstract(file, chat))
names(results) <- basename(abstract_files)
results
```

Even with `required = FALSE` I still saw hallucinated answers — a year that was simply wrong for one study. Switching to Claude Sonnet (a more capable model) fixed some but not all of these. This is a genuinely hard problem, not a one-line fix.

## Reflections

**Cost** is usually the least of your worries — this whole exercise, including testing, cost under 1c. Extracting from hundreds of methods sections is plausibly under $100. Cost only gets uncertain if you're iterating a lot on prompts, or feeding in full papers rather than abstracts.

**Getting clean text in** is the bigger challenge — downloading PDFs is slow, and paywalled HTML is often worse. HTML is preferable when you can get it, because tags already give you some structure.

**Validation matters more than ever** — if your review covers 1000 papers, plan to manually check a sample (say 100) and report the accuracy you found, the same as you would for any other measurement instrument.

**You still need to read the papers.** A lit review is more than extracted numbers — if you only use AI extraction you're vulnerable to the ['illusion of understanding'](https://www.nature.com/articles/s41586-024-07146-0) [@messeri_artificial_2024]. This tool is best suited to well-defined, narrow extraction tasks across consistently structured papers (e.g. pulling model types out of 500 species distribution model papers), not as a substitute for reading broadly.

::: {.callout-important title="Challenge" appearance="simple"}
Define a `paper_stats` object tailored to extracting the response variable, predictor variables, and statistical family used, from an abstract about an ecological count-data analysis (real or invented). Run it on two abstracts and manually check whether the extracted family (e.g. Poisson, negative binomial) is actually correct.
:::
